Polynomial Chaos for the Estimation of Manning's Friction

### 2013-11-13 paper/coefs.tex
minor edits and added class files

We seek to infer the Manning coefficient from water surface elevation
data measured at gauges as showm.
Bayesian inference can be applied directly to infer the uncertain parameters
using Equation~\ref{eq:post}. In our case, the observation data $\vec d$ 
are the water surface elevation at the gauges;
their \emph{GeoClaw} model predicted counterparts by $\vec G$.
The uncertain parameters $\vec \theta$ will be denoted by $\{N_i\}_{i=1}^3$ for the
three Manning Friction coefficients treated as the unknown parameters. 
In this case  Equation~\ref{eq:post} can be written as:
\begin{equation} 
p(N_1,N_2,N_3,\sigma^2 | \vec d) 
\propto \frac{1}{\sqrt{2 \pi \sigma^2}} 
 \prod_{i=1}^m  
\exp \left\lbrace \frac{-(d_i - G_i)^2}{2 \sigma^2} \right\rbrace
\ p(\sigma^2)p(N_1)p(N_2) p(N_3)
\label{eq:post_coef}
\end{equation}

To complete deinfe the posterior equation, we need to choose a proper prior that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for the model parameters is assumed as indicated:
\begin{equation} 
p(N_i) = \begin{cases}
		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i <  N_i \leq b_i ,  \\
		0 &\text{otherwise}  , 
\end{cases}
\end{equation}
where $ [a_i,b_i]$ denote the parameter ranges defined as.
Regarding the variance, the only information we know 
is that $\sigma^2$ is always positive.
We thus assume a Jeffreys prior \citep{sivia}, expressed as:
\begin{equation} 
p(\sigma^2) =  \begin{cases}
		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\
		0 &\text{otherwise}. 
		\end{cases}
\label{eq:var_pr}
\end{equation}

>>>>

We seek to infer the Manning coefficient from water surface elevation
data measured at different gauges as shown in Figure~\ref{fig:setup}.
Bayesian inference can be applied directly to infer the uncertain parameters
using Equation~\ref{eq:post}. In our case, the observation data $\vec d$ 
are the water surface elevation at the gauges;
their model predicted counterparts $\vec G$ is predicted by \emph{GeoClaw}.
The uncertain parameters $\vec \theta$ will be denoted by $\{N_i\}_{i=1}^3$ for the
three Manning Friction coefficients treated as the unknown parameters. 
In this case  Equation~\ref{eq:post} can be written as:
\begin{equation} 
p(N_1,N_2,N_3,\sigma^2 | \vec d) 
\propto \frac{1}{\sqrt{2 \pi \sigma^2}} 
 \prod_{i=1}^m  
\exp \left\lbrace \frac{-(d_i - G_i)^2}{2 \sigma^2} \right\rbrace
\ p(\sigma^2)p(N_1)p(N_2) p(N_3)
\label{eq:post_coef}
\end{equation}

To complete the definition of the posterior, we need to choose a proper prior that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for the model parameters is assumed as indicated:
\begin{equation} 
p(N_i) = \begin{cases}
		\displaystyle \frac{1}{b_i-a_i} &\text{for~} a_i <  N_i \leq b_i ,  \\
		0 &\text{otherwise}  , 
\end{cases}
\end{equation}
where $ [a_i,b_i]$ denote the parameter ranges defined as.
Regarding the variance, the only information we know 
is that $\sigma^2$ is always positive.
We thus assume a Jeffreys prior \citep{sivia}, expressed as:
\begin{equation} 
p(\sigma^2) =  \begin{cases}
		\displaystyle \frac{1}{\sigma^2} &\text{for~} \sigma^2 > 0,  \\
		0 &\text{otherwise}. 
		\end{cases}
\label{eq:var_pr}
\end{equation}

### 2013-11-13 paper/formulation.tex
minor edits and added class files

\section{Formulation}
\label{sec:formu}
We begin by introducing the essential building blocks of the present formulation: 
the Bayesian approach to inverse problems~\ref{sec:inference}
and the stochastic spectral methods for forward uncertainty propagation in ~\ref{sec:uqpce}
\input{inference}
\input{uqpce}

>>>>

\section{Formulation}
\label{sec:formu}
In this section, we introduce the methods employed in this work.
First, the Bayesian approach to inverse problems is described in Section~\ref{sec:inference};
its application to our problem of inferring the Manning Friction Coefficients in Section~\ref{sec:manning}.
Finally, in Section~\ref{sec:uqpce}, we introduce
a stochastic spectral method for forward uncertainty propagation and accelerating the Bayesian inference.

\input{inference}
\input{coefs}

\input{uqpce}

### 2013-11-13 paper/inference.tex
minor edits and added class files

Bayesian inference is a statistical approach to inverse problems
that has recently gained much interest in different applications including
ocean~\citep{Alexanderian2011a,Zedler2012,sraj:2013a}
climate~\citep{OlsonEtAl2012} and geophysical~\citep{Malinverno2002} modeling.
The key idea in this approach is to express all forms of uncertainty
in terms of probability. The first step in a Bayesian inference 
is to formulate the forward problem (response surface/model) using 
a suitable likelihood function and a product of conditional probability densities. 
We then formulate the prior distribution over the unknown parameters of the model
using our best belief. Lastly, given some observation data, Bayes rule 
is used to obtain a posterior distribution for these unknown
\citep{sivia}. We review this approach briefly below.

Let $\vec{d}=(d_1,...,d_m)^T$ be a vector of observation data and $\vec{\theta}=(\theta_1,...,\theta_n)^T$ be a vector of model parameters or inputs. We now consider a forward model $\vec G$ that predicts the data function of 
the parameters such that:
\begin{equation}
\vec d \approx \vec{G}( \vec \theta)
\end{equation}

>>>>

Bayesian inference is a statistical approach to inverse problems
that has recently gained much interest in different applications including
ocean~\citep{Alexanderian2011a,Zedler2012,sraj:2013a}
climate~\citep{OlsonEtAl2012} and geophysical~\citep{Malinverno2002} modeling.
The key idea in this approach is to express all forms of uncertainty
in terms of probability. The first step in Bayesian inference 
is to formulate the forward problem (response surface/model) using 
a suitable likelihood function and a product of conditional probability densities. 
The prior distributions over the unknown parameters of the model
are then formulated using our best belief. Lastly, given some observation data, Bayes rule 
is used to obtain a posterior distribution for these unknown parameters
\citep{sivia}. We review this approach briefly below.

Let $\vec{d}=(d_1,...,d_m)^T$ be a vector of observation data and $\vec{\theta}=(\theta_1,...,\theta_n)^T$ be a vector of model parameters or inputs. We consider a forward model $\vec G$ that predicts the data function of 
the parameters such that:
\begin{equation}
\vec d \approx \vec{G}( \vec \theta)
\end{equation}

### 2013-11-13 paper/uqpce.tex
minor edits and added class files

MCMC exploration of the 4D posterior (Equation~\ref{eq:post_coef}) requires repeated simulations (tens of thousands) of the forward GeoClaw model, once for each proposed move of the Markov chain. While a single \emph{Geoclaw} simulation
takes $\sim 15~$mins, depending on the details of the MCMC sampler, it is desirable to avoid repeated forward solutions altogether. An alternative is to construct a surrogate model that requires a much
smaller ensemble of GeoClaw runs, and that can be used instead
at a significantly reduced computational cost.  Here, we rely on
Polynomial Chaos expansions for accelerating Bayesian inference in this context 
by building a surrogate, which, in addition
also provide statistical properties,  such as the mean, variance and sensitivities,
efficiently.

@@

Let $U=U(\bm{x},t,\xxi)$ denote a quantity of 
interest (QoI) that is the output of certain model.
$U$ is normally function of space $\bm{x}$ and time $t$, and 
also on the vector of normalized uncertain model inputs $\xxi=(\xi_1,...,\xi_n)$ (called the germ). PC expresses $U$ in the form:
\begin{equation}
  U(\xbold,t,\xxi) \doteq \sum_{k = 0}^P U_k(\xbold,t) \Psi_k(\xxi)
\label{eq:stochseries}
\end{equation} 
where $U_k(\xbold,t)$ are the polynomial coefficients, and
$\Psi_k(\xxi)$ are functions that form an orthogonal basis of an underlying probability
space. The total number of terms in the PC expansion is
$P+1 = \frac{(d+p)! }{n!\ p!}$ where $n$ is the number of stochastic dimension and $p$ is the highest order
of polynomials. 
This series representation can be viewed as a spectral expansion
of $U$ along the stochastic dimensions.  The existence and convergence of this series
is asserted by the Cameron-Martin theorem
\citep{Cameron:1947}  whenever $U$ has finite variance. The series rate of convergence, and
hence the number of terms to retain, depends on the smoothness of
$U$ with respect to $\xxi$. The series converges spectrally fast with $P$
when $U$ is infinitely smooth; the convergence rate becomes algebraic
when $U$ has finite smoothness \citep{Canuto:2006}. 
%In practice the series convergence is monitored 
%via various error metrics; Section \ref{sec:results}\ref{sec:analysis} presents one possible error
%analysis for the present problem.

The choice of basis is dictated by the probability density
function of the stochastic variable $\xxi$, and which appears as a weight
function in the stochastic space's inner product:
\begin{equation}
 \left<\Psi_i,\Psi_j\right> = \int \Psi_i(\xxi) \;\Psi_j(\xxi) \; \rho(\xxi) \; \mbox{d}\xxi=\delta_{ij}\ave{\Psi_i^2}
\label{eq:inner}
\end{equation}
where $\delta_{ij}$ is the Kronecker delta.
For uniform
distributions the basis functions are hence scaled Legendre polynomials.
For multi-dimensional problems the basis functions are
tensor products of 1D basis functions. The identification of the inner product weight function
with the probability distribution of $\xxi$ simplifies the calculations of $U$'s statistical moments.
Noting that since $\Psi_0(\xxi)$ is a constant that can be normalized to satisfy 
$\left<\Psi_0,\Psi_0\right>=1$, the expectation and variance of $U$ can be computed as:
\begin{eqnarray}
 E[U]&=&\int U \, \rho(\xxi) \, \mbox{d}\xxi=\left< U,\Psi_0\right> = U_0  
 \label{eq:mean} \\
 E[(U-E[U])^2]&=&\int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi=\sum_{k=1}^P U_k^2
 \label{eq:sigma}
\left<\Psi_k,\Psi_k\right>
\end{eqnarray}

>>>>

MCMC exploration of the four-dimensional posterior (Equation~\ref{eq:post_coef}) requires repeated simulations (tens of thousands) of the forward GeoClaw model, once for each proposed move of the Markov chain. While a single \emph{Geoclaw} simulation
takes $\sim 15~$mins, depending on the details of the MCMC sampler, it is desirable to avoid repeated forward solutions altogether. An alternative is to construct a surrogate model that requires a much
smaller ensemble of GeoClaw runs, and that can be used instead
at a significantly reduced computational cost.  Here, we rely on
Polynomial Chaos expansions for accelerating Bayesian inference in this context 
by building a surrogate, which, in addition
also provide statistical properties,  such as the mean, variance and sensitivities,
efficiently.

@@

Let $U=U(\bm{x},t,\xxi)$ denote a quantity of 
interest (QoI) that is the output of certain model.
$U$ is normally function of space $\bm{x}$ and time $t$, and 
also on the vector of normalized uncertain model inputs $\xxi=(\xi_1,...,\xi_n)$ (called the germ). PC expresses $U$ in the form:
\begin{equation}
  U(\xbold,t,\xxi) \doteq \sum_{k = 0}^P U_k(\xbold,t) \Psi_k(\xxi)
\label{eq:stochseries}
\end{equation} 
where $U_k(\xbold,t)$ are the polynomial coefficients, and
$\Psi_k(\xxi)$ are functions that form an orthogonal basis of an underlying probability
space. The total number of terms in the PC expansion is
$P+1 = \frac{(d+p)! }{n!\ p!}$ where $n$ is the number of stochastic dimension and $p$ is the highest order
of polynomials. 
\comment{This series representation can be viewed as a spectral expansion
of $U$ along the stochastic dimensions.  The existence and convergence of this series
is asserted by the Cameron-Martin theorem
\citep{Cameron:1947}  whenever $U$ has finite variance. The series rate of convergence, and
hence the number of terms to retain, depends on the smoothness of
$U$ with respect to $\xxi$. The series converges spectrally fast with $P$
when $U$ is infinitely smooth; the convergence rate becomes algebraic
when $U$ has finite smoothness \citep{Canuto:2006}. }
%In practice the series convergence is monitored 
%via various error metrics; Section \ref{sec:results}\ref{sec:analysis} presents one possible error
%analysis for the present problem.

The choice of basis is dictated by the probability density
function of the stochastic variable $\xxi$, and which appears as a weight
function in the stochastic space's inner product:
\begin{equation}
 \left<\Psi_i,\Psi_j\right> = \int \Psi_i(\xxi) \;\Psi_j(\xxi) \; \rho(\xxi) \; \mbox{d}\xxi=\delta_{ij}\ave{\Psi_i^2}
\label{eq:inner}
\end{equation}
where $\delta_{ij}$ is the Kronecker delta.
For uniform
distributions the basis functions are hence scaled Legendre polynomials.
For multi-dimensional problems the basis functions are
tensor products of 1D basis functions. The identification of the inner product weight function
with the probability distribution of $\xxi$ simplifies the calculations of $U$'s statistical moments.

\subsubsection{Use of Polynomial Chaos}
Noting that since $\Psi_0(\xxi)$ is a constant that can be normalized to satisfy 
$\left<\Psi_0,\Psi_0\right>=1$, the expectation and variance of $U$ can be computed as:
\begin{eqnarray}
 E[U]&=&\int U \, \rho(\xxi) \, \mbox{d}\xxi=\left< U,\Psi_0\right> = U_0  
 \label{eq:mean} \\
 E[(U-E[U])^2]&=&\int (U-E[U])^2 \, \rho(\xxi) \, \mbox{d}\xxi=\sum_{k=1}^P U_k^2
 \label{eq:sigma}
\left<\Psi_k,\Psi_k\right>
\end{eqnarray}
